```{r options, include=FALSE}
opts_chunk$set(cache = TRUE)
```

Ohio Land Use and Environmental Variable Processing
========================================================

Jillian Deines  
March 2014  

This script summarizes land use and environmental variables by county for analyses. Updated from previous .r scripts.

#### Inputs:
* NLCD2006_OH.img - NLCD 2006 dataset clipped to Ohio in ArcGIS 
* OHsoil.shp - NRCS STATSGO soil file (attributes joined in ArcGIS)
* Environmental variable layers (described below)

#### Crunching:
* Summarizes land use by county
* Process land use (category combinations, NMDS)
* Rasterizes soil variables of interest
* Extracts mean environmental variable value by county
* Environmental variable plots

#### Outputs:
* OHCountyLandCover.csv - land use summary
* Soil layer rasters
  * OH_soil_awc.img = available water content
  * OH_soil_ph.img  = soil ph
  * OH_soil_slope.img = slope
* OH_EnviroVarCountySummary.csv - environmental variables summary

#### Next steps:
* Run analyses in OHAnalyses.Rmd

Working directory and packages
------------------------------
This assumes you have the file 'CleanCodeAndData' set as your working directory

```{r setup,message=FALSE}
library(rgdal)
library(raster)
library(RColorBrewer)
```

Land Use Summary by County
---------------------------
This section takes the National Land Cover Database (2006) raster for Ohio and extracts the percentages of each type of land use by county

NOTE: The overlay in this session takes a decent amount of time to run (1 hour +).

```{r landUseSummary, eval=FALSE}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# land cover raster
landUse <- raster("GISlayers/NLCD_OH/NLCD2006_OH.img")
ohio <- readOGR("GISlayers/Ohio","Ohio")
landUse2 <- mask(landUse, ohio, filename='GISlayers/NLCD_OH_mask.img')
# county polygons
county <- readOGR("GISlayers","OHCounty", verbose=F)

#---extract raster values by county--- 
ovRc <- extract(landUse,county)

# use table to summarize list ovRc (county)
tabCount <- lapply(ovRc,table)

# land class options
landclass<-c(11,21,22,23,24,31,41,42,43,52,71,81,82,90,95) 

# blank matrix with a column for each land class and a row for each polygon
landcoverC<-matrix(0,ncol=length(landclass),nrow=length(tabCount)) 

# fill matrix with land class percentages
for(i in 1:length(tabCount)){
  sc <- sum(tabCount[[i]])    	# total number of cells in each polygon
	landcoverC[i, landclass%in%names(tabCount[[i]])] <- as.numeric(tabCount[[i]]/sc)
}

#turn matrix into a data frame and add a column with total number of raster cells in each polygon
landcoverAC <- cbind(as.data.frame(landcoverC), numcells=sapply(tabCount,sum)) 
#add land cover class names to columns in data.frame
names(landcoverAC)[1:15]<-landclass     
landcoverAC <- cbind(Polygon = rownames(landcoverAC), landcoverAC)

# replace column names with actual land cover names
colnames(landcoverAC)[2:16] <- c("openWater", "developedOpen", "developedLow", "developedMedium", "developedHigh", "barren", "forestDec", "forestEver", "forestMix", "shrub", "grassland", "pasture", "crops", "wetlandsWoody", "wetlandsHerb")

# add county names to output
countyKey <- as.data.frame(county[,11:12])
landcover2 <- merge(landcoverAC,countyKey,by.x='Polygon', by.y='ID', sort=F)
landcover2 <- landcover2[,c(1,18,2:17)] # rearrange

write.csv(landcover2,"R output/OHCountyLandCover.csv", row.names=F)
```


Environmental Variables
-----------------------------
Process environmental variable GIS layers for OH and summarize these variables by county. Variables include:
* annual precipitation (PRISM climate normals)
* mean annual temperature (Worldclim)
* minimum January temperature (PRISM climate normals)
* maximum July temperature (PRISM climate normals)
* soil pH (STATSGO)
* soil available water content (STATSGO)
* slope (STATSGO)

### Rasterize soil data
OHsoil.shp represents NCRS STATSGO soil data for Ohio. I originally joined the soil variables and the shapefile in ArcGIS at earlier phases of this project, so I will not repeat that here.

```{r soilToRaster, eval=FALSE}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# soil data
soil <- readOGR("GISlayers","OHsoil")

# blank raster grid (5 arc second resolution)
r <- raster(ncol=516, nrow=429, crs="+proj=longlat +datum=NAD83")  # create blank raster file
extent(r) <- extent(soil)    					# give blank raster file same extent as the polygon file

# awc_r = available water content
soilwater <- rasterize(soil, r, 'awc_r', fun = mean, filename='GISlayers/OH_soil_awc.img')
# slope_r  = average slope
slope <- rasterize(soil, r, 'slope_r', fun = mean, filename='GISlayers/OH_soil_slope.img')
# ph1to1h2o_r = ph -> truncated to ph1to1h2o1 when read in
soilph <- rasterize(soil, r, 'ph1to1h2o1', fun = mean, filename='GISlayers/OH_soil_ph.img')
```

### Create function for extracting mean raster values by temperature

```{r extractFunction}
extractVar <- function(rasterfile, spdf) {
  # Extracts raster cell values by polygon and takes the mean.
  # INPUTS: 
    # raster = raster file path
    # polydir = directory containing polygon
    # polygon = polygon file name, no extension  
  # RETURNS: 
  
  ras  <- raster(rasterfile)
  
  # extract raster values by polygon (takes a minute to run)
  overList = extract(ras,spdf)
  
  # get mean raster value by polygon 
  meanVal <- sapply(overList,mean, na.rm=T)
  
  return(meanVal)
}

```

### Extract county mean values for variables
Additional data source information:
* Annual precipitation, Minimum January Temperature, and Maximum January Temperature: data comes from PRISM Climate Group, Oregon State University. 2004. Available online at http://prism.oregonstate.edu. Accessed October 2011. The data are normals for 1971-2000; 1981-2010 became available in 2012 but are not used here. Data have a 30 arc-second (800 m) grid resolution; this file was clipped to the PalEON domain in ArcGIS.
* Mean temperature: data comes from Worldclim, http://www.worldclim.org/current.Accessed fall 2012. Methodology for data here: http://www.worldclim.org/methods. This data is the average temperature for 1960-1990 (generally), .08333 grid cell size. Data is distributed tmean by month; in ArcGIS, we converted the GCS from WGS84 to NAD83, averaged all 12 months to get annual average, and clipped to the PalEON domain 
* Soil: statsgo (see above)

```{r extractValues, eval=FALSE}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# county boundary shapefile
county <- readOGR("GISlayers","OHCounty", verbose=F) 

#----Get county Means----#
precip <- 'GISlayers/ppt_annuPal.img'
meanPrecip <- extractVar(precip, county)
meanPrecip <- meanPrecip / 100

tmean <- 'GISlayers/tmean14Pal.img'
meanTemp <- extractVar(tmean, county)
meanTemp <- meanTemp / 10

tminjan <- "GISlayers/tminjanP.img"
meantmin <- extractVar(tminjan, county)
meantmin <- meantmin / 100

tmaxjul <- "GISlayers/tmaxjulP.img"
meantmax <- extractVar(tmaxjul, county)
meantmax <- meantmax / 100

soilh20 <- "GISlayers/OH_soil_awc.img"
meansoilwater <- extractVar(soilh20, county)

soilSlope <- 'GISlayers/OH_soil_slope.img'
meanslope <- extractVar(soilSlope, county)

soil_ph <- 'GISlayers/OH_soil_ph.img'
meansoilph <- extractVar(soil_ph, county)

OHenvar <- data.frame(countyID = county$ID,
                      countyName = county$CountyName, 
                      soilph = meansoilph, 
                      slope = meanslope, 
                      soilh2o = meansoilwater, 
                      tempann = meanTemp, 
                      pptann = meanPrecip, 
                      tminjan = meantmin, 
                      tmaxjul = meantmax)

write.csv(OHenvar, 'R output/OH_EnviroVarCountySummary.csv', row.names=F)
```

Plot Ohio Environmental Variables
----------------------------

```{r enviroVar_fig, fig.width=8,fig.height=4}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

landUse <- raster("GISlayers/NLCD_OH/NLCD2006_OH.img")
envar <- read.csv('R output/OH_EnviroVarCountySummary.csv')
county <- readOGR("GISlayers","OHCounty", verbose=F)
county.proj <- spTransform(county[,11:12], CRS("+init=epsg:32123"))

# add envar attribute data to county polygons
envarMap <- merge(county.proj, envar, by.x = "ID", by.y = "countyID", all.x = TRUE)

# store a bunch of spplots to combine below
slope <- spplot(envarMap[,5], col.regions=terrain.colors(16))
soilph <- spplot(envarMap[,4], col.regions=heat.colors(16))
soilh20 <- spplot(envarMap[,6], col.regions=rev(topo.colors(16)))
pptann <- spplot(envarMap[,8], col.regions = rev(topo.colors(16)))
tempann <- spplot(envarMap[,7], col.regions = rev(brewer.pal(10, "RdYlBu")), 
                  at=seq(17.4,20.3,((20.3-17.4)/10)))
tminjan <- spplot(envarMap[,9], col.regions = rev(brewer.pal(9, "Blues")), 
                  at=seq(-9.66, -5.86, ((9.66-5.86)/9)))
tmaxjul <- spplot(envarMap[,10], col.regions = brewer.pal(9, "OrRd"), 
                  at=seq(27.20, 30.35, ((30.35-27.20)/9)))
#plot(landUse)

# plot trellis objects in a matrix
plot(pptann, split=c(1,1,4,2), main="Annual Precip", more=TRUE)
plot(tempann, split=c(2,1,4,2), main="Mean Annual Temp", more=TRUE)
plot(tminjan, split=c(3,1,4,2), main="Min January Temp", more=TRUE)
plot(tmaxjul, split=c(4,1,4,2), main="Max July Temp", more=TRUE)
plot(soilph, split=c(1,2,4,2), main="Soil pH", more=TRUE)
plot(soilh20, split=c(2,2,4,2), main="Soil Water", more=TRUE)
plot(slope, split=c(3,2,4,2), main="Average Slope", more=FALSE)
#plot(landuse, split=c(4,2,4,2), more=FALSE)





```
