```{r options, include=FALSE}
opts_chunk$set(cache = TRUE)
```

Ohio Land Use and Environmental Variable Processing
========================================================

Jillian Deines  
March 2014  

This script summarizes land use and environmental variables by county for analyses. Updated from previous .r scripts.

#### Inputs:
* NLCD2006_OH.img - NLCD 2006 dataset clipped to Ohio in ArcGIS 
* OHsoil.shp - NRCS STATSGO soil file (attributes joined in ArcGIS)
* Environmental variable layers (described below)

#### Crunching:
* Summarizes land use by county
* Process land use (category combinations, NMDS)
* Rasterizes soil variables of interest
* Extracts mean environmental variable value by county
* Environmental variable plots

#### Outputs:
* OHCountyLandCover.csv - land use summary
* OHCountyLandCover_aggregated_nmds.csv - includes nmds values
* Soil layer rasters
  * OH_soil_awc.img = available water content
  * OH_soil_ph.img  = soil ph
  * OH_soil_slope.img = slope
* OH_EnviroVarCountySummary.csv - environmental variables summary

#### Next steps:
* Run analyses in OHAnalyses.Rmd

Working directory and packages
------------------------------
This assumes you have the file 'CleanCodeAndData' set as your working directory

```{r setup,message=FALSE}
library(rgdal)
library(raster)
library(ecodist) 
library(RColorBrewer)
```

Land Use Summary by County
---------------------------

### Land Use Summary
This section takes the National Land Cover Database (2006) raster for Ohio and extracts the percentages of each type of land use by county

NOTE: The overlay in this session takes a decent amount of time to run (1 hour +). Output has been written to .csv.

```{r landUseSummary, eval=FALSE}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# land cover raster
landUse <- raster("GISlayers/NLCD_OH/NLCD2006_OH.img")
ohio <- readOGR("GISlayers/Ohio","Ohio")
landUse2 <- mask(landUse, ohio, filename='GISlayers/NLCD_OH_mask.img')
# county polygons
county <- readOGR("GISlayers","OHCounty", verbose=F)

#---extract raster values by county--- 
ovRc <- extract(landUse,county)

# use table to summarize list ovRc (county)
tabCount <- lapply(ovRc,table)

# land class options
landclass<-c(11,21,22,23,24,31,41,42,43,52,71,81,82,90,95) 

# blank matrix with a column for each land class and a row for each polygon
landcoverC<-matrix(0,ncol=length(landclass),nrow=length(tabCount)) 

# fill matrix with land class percentages
for(i in 1:length(tabCount)){
  sc <- sum(tabCount[[i]])    	# total number of cells in each polygon
	landcoverC[i, landclass%in%names(tabCount[[i]])] <- as.numeric(tabCount[[i]]/sc)
}

#turn matrix into a data frame and add a column with total number of raster cells in each polygon
landcoverAC <- cbind(as.data.frame(landcoverC), numcells=sapply(tabCount,sum)) 
#add land cover class names to columns in data.frame
names(landcoverAC)[1:15]<-landclass     
landcoverAC <- cbind(Polygon = rownames(landcoverAC), landcoverAC)

# replace column names with actual land cover names
colnames(landcoverAC)[2:16] <- c("openWater", "developedOpen", "developedLow", "developedMedium", "developedHigh", "barren", "forestDec", "forestEver", "forestMix", "shrub", "grassland", "pasture", "crops", "wetlandsWoody", "wetlandsHerb")

# add county names to output
countyKey <- as.data.frame(county[,11:12])
landcover2 <- merge(landcoverAC,countyKey,by.x='Polygon', by.y='ID', sort=F)
landcover2 <- landcover2[,c(1,18,2:17)] # rearrange

write.csv(landcover2,"R output/OHCountyLandCover.csv", row.names=F)
```


### Aggregate Land Use Data and NMDS
Combines similar land use classes and collapses county land use composition into one variable along the 'principle component' for subsequent analyses. NMDS is run for all Ohio counties, as well as on the subset of 27 counties with PLS data.

Chunk already run with output saved.

```{r landUse_agg_NMDS, eval=FALSE}
# load county land use summary
OHlc<-read.csv("R Output/OHCountyLandCover.csv")

# clean data: combine similar land use classes 
OHlc1 <- cbind(OHlc[,c('Polygon','CountyName','openWater','developedOpen')],
               "developed"=rowSums(OHlc[,c('developedLow','developedMedium',
                                           'developedHigh')]),
                "barren" = OHlc$barren,
                "forest" = rowSums(OHlc[,c('forestDec','forestEver','forestMix')]),
                OHlc[,c('shrub','grassland','pasture','crops')],  
                "wetlands"=rowSums(OHlc[,c('wetlandsWoody','wetlandsHerb')]),
                "numcells" = OHlc$numcells)   

# remove county polygons that have no overlap with land use data (ie, parts of Ottawa county that are islands in Lake Michigan)
lcNmds <- OHlc1[OHlc1$numcells != 0,]

# subset to 27 data counties, add coords
centroids <- read.csv("R output/OH_county_centroids_projected.csv") 
lcNmds27 <- merge(centroids, lcNmds, by.x = "ID", by.y = "Polygon", all.y = FALSE)

# arcsine squareroot transform proportional data 
lcNmds[,3:12] <- asin(sqrt(lcNmds[,3:12]))     
lcNmds27[,5:14] <- asin(sqrt(lcNmds27[,5:14]))  

# calculate distance matrix on land use data columns
library(ecodist) 
lcDist <- ecodist::distance(lcNmds[,3:12], method = "euclidean")
lcDist27 <- ecodist::distance(lcNmds27[,5:14], method = "euclidean")

# Run NMDS with only 1 dimension to generate 1 land use value for each county for plotting/correlations:
nmdsAll <- nmds(lcDist, maxdim = 1)
nmds27 <- nmds(lcDist27, maxdim = 1)

# View the results for each interation/configurations
nmdsAll$stress	# can be thought of as 1-r2
nmdsAll$r2		# Percent variation each configuration explains

nmds27$r2

# get minimum stress values for each county 
nmdsMin <- nmds.min(nmdsAll)  
nmdsMin$ID <- lcNmds$Polygon
colnames(nmdsMin) <- c('nmdsAll','ID')

nmdsMin27 <- nmds.min(nmds27)  
nmdsMin27$ID <- centroids[,1] 
colnames(nmdsMin27) <- c('nmds27','ID')

# add nmds output to csv
lcNmdsOut <- merge(OHlc1,nmdsMin, by.x='Polygon', by.y='ID',all=T)
lcNmdsOut2 <- merge(lcNmdsOut,nmdsMin27, by.x='Polygon', by.y='ID',all=T)

# visualize nmds output
county <- readOGR("GISlayers","OHCounty", verbose=F)
countyTest <- merge(county, lcNmdsOut2[,c(1,14:15)], by.x='ID',by.y = 'Polygon')
spplot(countyTest,c('nmdsAll','nmds27'), col.regions=colorRampPalette(c('darkmagenta','gray88', 'forestgreen'))(20))

write.csv(lcNmdsOut2,"R Output/OHCountyLandCover_aggregated_nmds.csv",row.names=F)
```


Environmental Variables
-----------------------------
Process environmental variable GIS layers for OH and summarize these variables by county. Variables include:
* annual precipitation (PRISM climate normals)
* mean annual temperature (Worldclim)
* minimum January temperature (PRISM climate normals)
* maximum July temperature (PRISM climate normals)
* soil pH (STATSGO)
* soil available water content (STATSGO)
* slope (STATSGO)

### Rasterize soil data
OHsoil.shp represents NCRS STATSGO soil data for Ohio. I originally joined the soil variables and the shapefile in ArcGIS at earlier phases of this project, so I will not repeat that here.

```{r soilToRaster, eval=FALSE}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# soil data
soil <- readOGR("GISlayers","OHsoil")

# blank raster grid (5 arc second resolution)
r <- raster(ncol=516, nrow=429, crs="+proj=longlat +datum=NAD83")  # create blank raster file
extent(r) <- extent(soil)    					# give blank raster file same extent as the polygon file

# awc_r = available water content
soilwater <- rasterize(soil, r, 'awc_r', fun = mean, filename='GISlayers/OH_soil_awc.img')
# slope_r  = average slope
slope <- rasterize(soil, r, 'slope_r', fun = mean, filename='GISlayers/OH_soil_slope.img')
# ph1to1h2o_r = ph -> truncated to ph1to1h2o1 when read in
soilph <- rasterize(soil, r, 'ph1to1h2o1', fun = mean, filename='GISlayers/OH_soil_ph.img')
```

### Create function for extracting mean raster values by county

```{r extractFunction}
extractVar <- function(rasterfile, spdf) {
  # Extracts raster cell values by polygon and takes the mean.
  # INPUTS: 
    # raster = raster file path
    # polydir = directory containing polygon
    # polygon = polygon file name, no extension  
  # RETURNS: 
  
  ras  <- raster(rasterfile)
  
  # extract raster values by polygon (takes a minute to run)
  overList = extract(ras,spdf)
  
  # get mean raster value by polygon 
  meanVal <- sapply(overList,mean, na.rm=T)
  
  return(meanVal)
}

```

### Extract county mean values for variables
Additional data source information:
* Annual precipitation, Minimum January Temperature, and Maximum January Temperature: data comes from PRISM Climate Group, Oregon State University. 2004. Available online at http://prism.oregonstate.edu. Accessed October 2011. The data are normals for 1971-2000; 1981-2010 became available in 2012 but are not used here. Data have a 30 arc-second (800 m) grid resolution; this file was clipped to the PalEON domain in ArcGIS.
* Mean temperature: data comes from Worldclim, http://www.worldclim.org/current.Accessed fall 2012. Methodology for data here: http://www.worldclim.org/methods. This data is the average temperature for 1960-1990 (generally), .08333 grid cell size. Data is distributed tmean by month; in ArcGIS, we converted the GCS from WGS84 to NAD83, averaged all 12 months to get annual average, and clipped to the PalEON domain 
* Soil: statsgo (see above)

```{r extractValues, eval=FALSE}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# county boundary shapefile
county <- readOGR("GISlayers","OHCounty", verbose=F) 

#----Get county Means----#
precip <- 'GISlayers/ppt_annuPal.img'
meanPrecip <- extractVar(precip, county)
meanPrecip <- meanPrecip / 100

tmean <- 'GISlayers/tmean14Pal.img'
meanTemp <- extractVar(tmean, county)
meanTemp <- meanTemp / 10

tminjan <- "GISlayers/tminjanP.img"
meantmin <- extractVar(tminjan, county)
meantmin <- meantmin / 100

tmaxjul <- "GISlayers/tmaxjulP.img"
meantmax <- extractVar(tmaxjul, county)
meantmax <- meantmax / 100

soilh20 <- "GISlayers/OH_soil_awc.img"
meansoilwater <- extractVar(soilh20, county)

soilSlope <- 'GISlayers/OH_soil_slope.img'
meanslope <- extractVar(soilSlope, county)

soil_ph <- 'GISlayers/OH_soil_ph.img'
meansoilph <- extractVar(soil_ph, county)

OHenvar <- data.frame(countyID = county$ID,
                      countyName = county$CountyName, 
                      soilph = meansoilph, 
                      slope = meanslope, 
                      soilh2o = meansoilwater, 
                      tempann = meanTemp, 
                      pptann = meanPrecip, 
                      tminjan = meantmin, 
                      tmaxjul = meantmax)

write.csv(OHenvar, 'R output/OH_EnviroVarCountySummary.csv', row.names=F)
```

Plot Ohio Environmental Variables
----------------------------

```{r enviroVar_fig, fig.width=8,fig.height=4}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

landUse <- raster("GISlayers/NLCD_OH/NLCD2006_OH.img")
envar <- read.csv('R output/OH_EnviroVarCountySummary.csv')
county <- readOGR("GISlayers","OHCounty", verbose=F)
county.proj <- spTransform(county[,11:12], CRS("+init=epsg:32123"))

# add envar attribute data to county polygons
envarMap <- merge(county.proj, envar, by.x = "ID", by.y = "countyID", all.x = TRUE)

# store a bunch of spplots to combine below
slope <- spplot(envarMap[,5], col.regions=terrain.colors(16))
soilph <- spplot(envarMap[,4], col.regions=heat.colors(16))
soilh20 <- spplot(envarMap[,6], col.regions=rev(topo.colors(16)))
pptann <- spplot(envarMap[,8], col.regions = rev(topo.colors(16)))
tempann <- spplot(envarMap[,7], col.regions = rev(brewer.pal(10, "RdYlBu")), 
                  at=seq(17.4,20.3,((20.3-17.4)/10)))
tminjan <- spplot(envarMap[,9], col.regions = rev(brewer.pal(9, "Blues")), 
                  at=seq(-9.66, -5.86, ((9.66-5.86)/9)))
tmaxjul <- spplot(envarMap[,10], col.regions = brewer.pal(9, "OrRd"), 
                  at=seq(27.20, 30.35, ((30.35-27.20)/9)))
#plot(landUse)

# plot trellis objects in a matrix
plot(pptann, split=c(1,1,4,2), main="Annual Precip", more=TRUE)
plot(tempann, split=c(2,1,4,2), main="Mean Annual Temp", more=TRUE)
plot(tminjan, split=c(3,1,4,2), main="Min January Temp", more=TRUE)
plot(tmaxjul, split=c(4,1,4,2), main="Max July Temp", more=TRUE)
plot(soilph, split=c(1,2,4,2), main="Soil pH", more=TRUE)
plot(soilh20, split=c(2,2,4,2), main="Soil Water", more=TRUE)
plot(slope, split=c(3,2,4,2), main="Average Slope", more=FALSE)
#plot(landuse, split=c(4,2,4,2), more=FALSE)





```
