
```{r options, include=FALSE}
opts_chunk$set(cache = TRUE)
```

Ohio Analyses
========================================================

Jillian Deines  
March 2014

This script analyses correlations among Ohio vegetation (historic and modern) and covariates (land use and environmental variables).

Analyses aim to describe presettlement vegetation in Ohio and examine relationships among modern vegetation (using FIA data), historic vegetation, land use, and environmental variables.

#### Depends:
This script relies on output from the following scripts
* PLS_FIA_data_crunching.Rmd
* OH_Land_Use_Enviro_Var_processing.Rmd

#### Inputs:
* OHcountyLandCover.csv
* OH_EnviroVarCountySummary.csv
* OH_county_centroids_latlong.csv 
* OHCompCleanPLS_percent.csv
* OHCompCleanFIA_percent.csv

#### Crunching:
* Aggregate land use classes
* Basic correlation matrices between county composition (FIA/PLS) and 1) environmental variables and 2) land use
* Mantel tests: aggregated environmental and land use dissimilarity matrices
* NMDS on county composition
* Composition statistics by county/time period

#### Outputs:
* MantelResults.csv
* Correlation plots
* NMDS of composition for each time period (plots)

Working directory and packages
------------------------------
This assumes you have the file 'CleanCodeAndData' set as your working directory

```{r setup,message=FALSE}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')
library(ecodist)
library(corrplot)
library(RColorBrewer)
```


Correlations
--------------------------
PLS, FIA, environmental variables, and land use summaries are analyzed for correlations.  
* Proportional species composition data is arc sine square root transformed

**Note:** In previous versions of this script prior to Jill's pre-publication cleaning effort, there was a typo in the land use/composition correlations where the correlation between land use and PLS composition was stored as the variable 'fia.land', and vice versa. **Thus the patterns between land use and composition previously described in Dan's thesis version are reversed**. The number of the correlations differs due to data cleaning, but the pattern of strong/weak and positive/negative correlations is reversed.

These are not currently in the manuscript.

```{r Correlations}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# read in data
pls <- read.csv("R output/OHCompCleanPLS_percent.csv")
fia <- read.csv("R Output/OHCompCleanFIA_percent.csv")
envar <- read.csv("R output/OH_EnviroVarCountySummary.csv")
landuse <- read.csv("R Output/OHCountyLandCover_aggregated_nmds.csv")
centroids <- read.csv("R output/OH_county_centroids_latlong.csv") 

# subset enviro and land use to 27 data counties, add coords
envar2   <- merge(centroids, envar, by.x = "ID", by.y = "countyID", all.y = FALSE)
landuse2 <- merge(centroids, landuse, by.x = "ID", by.y = "Polygon", all.y = FALSE)

# arcsine squareroot transform proportional data (land use, pls, fia)
pls[,3:17]     <- asin(sqrt(pls[,3:17]))		
fia[,3:17]     <- asin(sqrt(fia[,3:17]))	
landuse2[,5:14] <- asin(sqrt(landuse2[,5:14])) 

# correlations: composition vs enviro, landuse 
plsEnvCor <-cor(envar2[,5:11], pls[,3:17])	
fiaEnvCor <-cor(envar2[,5:11], fia[,3:17])	

#plsLandCor <- cor(landuse2[,5:14], pls[,3:17])	
fiaLandCor <- cor(landuse2[,5:14], fia[,3:17])	
```

### Correlation Plots
Needed: Nicer axis labels, significance indication?
```{r CorrPlots, fig.width=9,fig.height=5}
#pls/environmental variables
corrplot(plsEnvCor, method="color", addCoef.col="black", tl.col="black", 
  col=colorRampPalette(c("darkmagenta", "white", "darkgreen"))(200))
#fia/enviro
corrplot(fiaEnvCor, method="color", addCoef.col="black", tl.col="black", 
  col=colorRampPalette(c("darkmagenta", "white", "darkgreen"))(200))

# #pls/land use
# corrplot(plsLandCor, method="color", addCoef.col="black", title="PLS-Land Use", tl.col="black", 
#   col=colorRampPalette(c("darkmagenta", "white", "darkorange2"))(200))
#fia/land use
corrplot(fiaLandCor, method="color", addCoef.col="black", tl.col="black", col=colorRampPalette(c("darkmagenta", "white", "darkorange2"))(200))
```


Mantel Tests
---------------------------

### Load Data and Get Distance Matrices

```{r Mantels_loadData, eval=FALSE}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

pls <- read.csv("R output/OHCompCleanPLS_percent.csv")
fia <- read.csv("R Output/OHCompCleanFIA_percent.csv")
envar <- read.csv("R output/OH_EnviroVarCountySummary.csv")
landuse <- read.csv("R Output/OHCountyLandCover_aggregated_nmds.csv")
centroids <- read.csv("R output/OH_county_centroids_latlong.csv") 

# subset enviro and land use to 27 data counties, add coords
envar2   <- merge(centroids, envar, by.x = "ID", by.y = "countyID", 
                  all.y = FALSE)
landuse2 <- merge(centroids, landuse, by.x = "ID", by.y = "Polygon", 
                  all.y = FALSE)

# arcsine squareroot transform proportional data (land use, pls, fia)
pls[,3:17] <- asin(sqrt(pls[,3:17]))  	
fia[,3:17] <- asin(sqrt(fia[,3:17]))	
landuse2[,5:14] <- asin(sqrt(landuse2[,5:14])) 
```

### Big Picture: Enviroment/climate vs Land Use
Reference: Goslee & Urban 2007  
Dissimilarity matrices for composition used the Bray-Curtis distance methoc, as recommended by Goslee & Urban.

First, I test for spatial autocorrelation in composition data; since this is positive, I use partial Mantel tests to test relationships of composition and land use/enviro

```{r Mantel1, eval=FALSE}
# composition distance matrices
FIAdist <- ecodist::distance(fia[,3:17], method= "bray-curtis") 
PLSdist <- ecodist::distance(pls[,3:17], method= "bray-curtis")

# ----- spatial autocoreelation -----------
spatialDist<- ecodist::distance(centroids[,c('X','Y')], method = "euclidean")

# output dataframe
MantelResults <- data.frame(variable = rep('tbd',30), dataset = NA, 
                            mantelr = NA, pval = NA)
MantelResults$variable <- as.character(MantelResults$variable)

# composition and space
MantelResults$variable[1:2] <- 'spatialAuto'
MantelResults$dataset[1:2] <- c('PLS','FIA')
MantelResults[1,3:4] <- mantel(PLSdist ~ spatialDist)[c('mantelr','pval1')]
MantelResults[2,3:4] <- mantel(FIAdist ~ spatialDist)[c('mantelr','pval1')]
# yes, data is spatially correlated

# ----- total enviro and land use ------------
luDist <- ecodist::distance(landuse2[,5:14], method = "euclidean")
envirDist <- ecodist::distance(envar2[,5:11], method = "euclidean")

# composition vs environment, given effects of space
MantelResults$variable[3:4] <- 'env_space'
MantelResults$dataset[3:4] <- c('PLS','FIA')
MantelResults[3,3:4] <- mantel(PLSdist ~ envirDist + spatialDist)[c('mantelr','pval1')]
MantelResults[4,3:4] <- mantel(FIAdist ~ envirDist + spatialDist)[c('mantelr','pval1')]

# land use given effects of space
MantelResults$variable[5:6] <- 'lu_space'
MantelResults$dataset[5:6] <- c('PLS','FIA')
MantelResults[5,3:4] <- mantel(PLSdist ~ luDist + spatialDist)[c('mantelr','pval1')]
MantelResults[6,3:4] <- mantel(FIAdist ~ luDist + spatialDist)[c('mantelr','pval1')]

# display results
head(MantelResults)
```

The Big Picture results indicate that variation in PLS and environmental dissimilarities matrices are significantly correlated. FIA and environment are not significantly correlated, but FIA is significantly correlated with modern land use. 

### Individual Environmental Variables
For both PLS and FIA data (but perhaps FIA data is inappropriate since the group test, FIA ~ enviro, was insignificant). For Dan's project, we ran the partial Mantel test with only space as the additional variables; the Git commit of this script on May 12, 2014 replicates Dan's results (except Dan's report does not show significance in FIA soil water and annual ppt, which mine do - we get the same mantel r, so I think Dan forgot to label the significant tests for FIA/environment), which are also reported in his final paper.

Here, I fully cross the partial Mantels, adding all other variables tested in addition to space. This generally changes the results as follows: 1) previously non-significant variables have much lower mantel r's and remain insignificant; and 2) previously significant variables are still significant (annppt p value lowered...), and their mantel r's are similar but a bit lower. **For FIA data, however, these changes had a larger effect: Annual precip became insignificant, and minimum temperature became significant. Weird.**


```{r Mantel_enviro, eval=FALSE}
# distance matrices for individual environmental variables examined
soilphDist  <- ecodist::distance(envar2[,'soilph'], method = "euclidean")
slopeDist   <- ecodist::distance(envar2[,'slope'], method = "euclidean")
soilh20Dist <- ecodist::distance(envar2[,'soilh2o'], method = "euclidean")
tempannDist <- ecodist::distance(envar2[,'tempann'], method = "euclidean") 
pptannDist  <- ecodist::distance(envar2[,'pptann'], method = "euclidean")
tminjanDist <- ecodist::distance(envar2[,'tminjan'], method = "euclidean")
tmaxjulDist <- ecodist::distance(envar2[,'tmaxjul'], method = "euclidean")

# add variables to output df
MantelResults$variable[7:20] <- rep(c('soilph','slope','soilh2o','tempann','pptann','tminjan','tmaxjul'),2)

# PLS tests
MantelResults$dataset[7:13] <- 'PLS'

MantelResults[7,3:4] <- mantel(PLSdist ~ soilphDist + spatialDist + slopeDist + soilh20Dist + tempannDist + pptannDist + tminjanDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[8,3:4] <- mantel(PLSdist ~ slopeDist + spatialDist + soilphDist + soilh20Dist + tempannDist + pptannDist + tminjanDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[9,3:4] <- mantel(PLSdist ~ soilh20Dist + spatialDist + soilphDist + slopeDist + tempannDist + pptannDist + tminjanDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[10,3:4] <- mantel(PLSdist ~ tempannDist + spatialDist + soilphDist + slopeDist + soilh20Dist + pptannDist + tminjanDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[11,3:4] <- mantel(PLSdist ~ pptannDist + spatialDist + soilphDist + slopeDist + soilh20Dist + tempannDist + tminjanDist + tmaxjulDist )[c('mantelr','pval1')]
MantelResults[12,3:4] <- mantel(PLSdist ~ tminjanDist + spatialDist + soilphDist + slopeDist + soilh20Dist + tempannDist + pptannDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[13,3:4] <- mantel(PLSdist ~ tmaxjulDist + spatialDist + soilphDist + slopeDist + soilh20Dist + tempannDist + pptannDist + tminjanDist)[c('mantelr','pval1')]

# FIA tests
MantelResults$dataset[14:20] <- 'FIA'

MantelResults[14,3:4] <- mantel(FIAdist ~ soilphDist + spatialDist + slopeDist + soilh20Dist + tempannDist + pptannDist + tminjanDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[15,3:4] <- mantel(FIAdist ~ slopeDist + spatialDist + soilphDist + soilh20Dist + tempannDist + pptannDist + tminjanDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[16,3:4] <- mantel(FIAdist ~ soilh20Dist + spatialDist + soilphDist + slopeDist + tempannDist + pptannDist + tminjanDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[17,3:4] <- mantel(FIAdist ~ tempannDist + spatialDist + soilphDist + slopeDist + soilh20Dist + pptannDist + tminjanDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[18,3:4] <- mantel(FIAdist ~ pptannDist + spatialDist + soilphDist + slopeDist + soilh20Dist + tempannDist + tminjanDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[19,3:4] <- mantel(FIAdist ~ tminjanDist + spatialDist + soilphDist + slopeDist + soilh20Dist + tempannDist + pptannDist + tmaxjulDist)[c('mantelr','pval1')]
MantelResults[20,3:4] <- mantel(FIAdist ~ tmaxjulDist + spatialDist + soilphDist + slopeDist + soilh20Dist + tempannDist + pptannDist + tminjanDist)[c('mantelr','pval1')]

# display results
MantelResults[7:20,]
```

### Individual Land Use Variables
For FIA data; fully adding variables to partial Mantel test resulted in forest land use no longer being significant.


```{r Mantel_lu, eval=FALSE}
# distance matrices for individual land use variables
Openh2odist <- ecodist::distance(landuse2[,'openWater'], method = "euclidean")
DevOpendist <- ecodist::distance(landuse2[,'developedOpen'], method = "euclidean")
Devdist     <- ecodist::distance(landuse2[,'developed'], method = "euclidean")
Barrendist  <- ecodist::distance(landuse2[,'barren'], method = "euclidean")
Forestdist  <- ecodist::distance(landuse2[,'forest'], method = "euclidean")
Shrubdist   <- ecodist::distance(landuse2[,'shrub'], method = "euclidean")
Grassdist   <- ecodist::distance(landuse2[,'grassland'], method = "euclidean")
Pasturedist <- ecodist::distance(landuse2[,'pasture'], method = "euclidean")
Cropsdist   <- ecodist::distance(landuse2[,'crops'], method = "euclidean")
Wetlanddist <- ecodist::distance(landuse2[,'wetlands'], method = "euclidean")

# add variables to output df
MantelResults$variable[21:30] <- c('water','open developed','developed','barren','forest','shrubland','grassland','pasture','crops','wetlands')

#partial mantel test with FIA data: land use
MantelResults$dataset[21:30] <- 'FIA'

MantelResults[21,3:4] <- mantel(FIAdist ~ Openh2odist + spatialDist + DevOpendist + Devdist + Barrendist + Forestdist + Shrubdist + Grassdist + Pasturedist + Cropsdist + Wetlanddist)[c('mantelr','pval1')]
MantelResults[22,3:4] <- mantel(FIAdist ~ DevOpendist + spatialDist + Openh2odist + Devdist + Barrendist + Forestdist + Shrubdist + Grassdist + Pasturedist + Cropsdist + Wetlanddist)[c('mantelr','pval1')]
MantelResults[23,3:4] <- mantel(FIAdist ~ Devdist + spatialDist + Openh2odist + DevOpendist + Barrendist + Forestdist + Shrubdist + Grassdist + Pasturedist + Cropsdist + Wetlanddist)[c('mantelr','pval1')]
MantelResults[24,3:4] <- mantel(FIAdist ~ Barrendist + spatialDist + Openh2odist + DevOpendist + Devdist + Forestdist + Shrubdist + Grassdist + Pasturedist + Cropsdist + Wetlanddist)[c('mantelr','pval1')]
MantelResults[25,3:4] <- mantel(FIAdist ~ Forestdist + spatialDist + Openh2odist + DevOpendist + Devdist + Barrendist + Shrubdist + Grassdist + Pasturedist + Cropsdist + Wetlanddist)[c('mantelr','pval1')]
MantelResults[26,3:4] <- mantel(FIAdist ~ Shrubdist + spatialDist + Openh2odist + DevOpendist + Devdist + Barrendist + Forestdist + Grassdist + Pasturedist + Cropsdist + Wetlanddist)[c('mantelr','pval1')]
MantelResults[27,3:4] <- mantel(FIAdist ~ Grassdist + spatialDist + Openh2odist + DevOpendist + Devdist + Barrendist + Forestdist + Shrubdist + Pasturedist + Cropsdist + Wetlanddist)[c('mantelr','pval1')]
MantelResults[28,3:4] <- mantel(FIAdist ~ Pasturedist + spatialDist + Openh2odist + DevOpendist + Devdist + Barrendist + Forestdist + Shrubdist + Grassdist + Cropsdist + Wetlanddist)[c('mantelr','pval1')]
MantelResults[29,3:4] <- mantel(FIAdist ~ Cropsdist + spatialDist + Openh2odist + DevOpendist + Devdist + Barrendist + Forestdist + Shrubdist + Grassdist + Pasturedist + Wetlanddist)[c('mantelr','pval1')]
MantelResults[30,3:4] <- mantel(FIAdist ~ Wetlanddist + spatialDist + Openh2odist + DevOpendist + Devdist + Barrendist + Forestdist + Shrubdist + Grassdist + Pasturedist + Cropsdist)[c('mantelr','pval1')]

# display
MantelResults[21:30,]

# write results
write.csv(MantelResults,"R Output/MantelResults.csv",row.names=F)
```

### Visualize Mantel Results in Markdown Output
```{r MantelOutput}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')
MantelResults <- read.csv("R Output/MantelResults.csv")

# group tests
head(MantelResults)

# enviro variables
MantelResults[7:20,]

# land use variables
MantelResults[21:30,]
```


Composition NMDS
----------------------------
We ran an ordination using non-metric multidimensional scaling to assess how composition differed by county in the two time periods.

```{r NMDS, eval=FALSE}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

pls <- read.csv("R output/OHCompCleanPLS_percent.csv")
fia <- read.csv("R Output/OHCompCleanFIA_percent.csv")

# arcsine squareroot transform proportional data (land use, pls, fia)
pls[,3:17] <- asin(sqrt(pls[,3:17]))    
fia[,3:17] <- asin(sqrt(fia[,3:17]))	

# composition distance matrices
FIAdist <- ecodist::distance(fia[,3:17], method= "bray-curtis") 
PLSdist <- ecodist::distance(pls[,3:17], method= "bray-curtis")

#run NMDS
PLSnmds <-nmds(PLSdist, maxdim=2)
FIAnmds <-nmds(FIAdist, maxdim=2)

# get the NMDS scores for each county for the configuration which had the minimum stress values
PLS.nmds.min <- nmds.min(PLSnmds)
colnames(PLS.nmds.min) <- c('NMDS 1','NMDS 2')
FIA.nmds.min <- nmds.min(FIAnmds)
colnames(FIA.nmds.min) <- c('NMDS 1','NMDS 2')

# calculate species vectors
PLS.nmds.vf<- vf(PLS.nmds.min, pls[,3:17])
FIA.nmds.vf<- vf(FIA.nmds.min, fia[,3:17]) 

# save these objects for plotting (to "freeze" stochasticity)
save(pls, fia, PLS.nmds.min, FIA.nmds.min, PLS.nmds.vf, FIA.nmds.vf, PLSnmds, FIAnmds, file='R Output/NMDS.RData')
```

```{r nmdsPlot, fig.width=9, fig.height=5}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')
load('R Output/NMDS.RData')

# assign county groups (1 = nw, 2 = se)
pls$region <- c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,2,1,2,2,1,2,2,2,2,2,2,2,2)
pls$color <- apply(as.array(pls$region),1,function (x) {
  if (x == 1) out <- 'gray10'
  else out <- 'gray60'
  return(out) })
pls$pch <- apply(as.array(pls$region),1,function (x) {
  if (x == 1) out <- 3
  else out <- 25
  return(out) })

names(PLS.nmds.min) <- c('Axis 1', 'Axis 2')
names(FIA.nmds.min) <- c('Axis 1', 'Axis 2')

#png("exportedFigs/ordination_bw.png", type = "cairo", units = "in", width = 7, height = 4, res = 2000)

# Plot NMDS min stress result - base plot
par(mfrow=c(1,2))

yrange <- c(-.45,0.58)
xrange <- c(-.45, 0.4)

par(mar=c(5,4,2,1))
plot(PLS.nmds.min, main='Settlement-Era', col=pls$color, pch=19, xlim=xrange, 
     cex=.8,ylim=yrange)
plot(PLS.nmds.vf,pch=1)
legend('topleft', legend=paste0('r^2 = ',round(max(PLSnmds$r2),2)), bty='n')  

par(mar=c(5,4,2,1))
plot(FIA.nmds.min, main='Modern', col=pls$color, pch=19, xlim=xrange, 
     cex=.8,ylim=yrange)
plot(FIA.nmds.vf)
legend('topleft', legend=paste0('r^2 = ',round(max(FIAnmds$r2),2)), bty='n') 

# empty plot over full device
par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), mar=c(0, 0, 0, 0), new=TRUE)
plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
# add legend (for example)
legend("bottom", legend=c('southeast counties','northwest counties'), pch=19, bty='n', col=c('gray60','gray10'))

## no arrow vectors
# extract the [x,y] coordinates from the PLS.nmds.vf object
#plot(PLS.nmds.vf[,1],PLS.nmds.vf[,2],pch=rownames(PLS.nmds.vf),add=T)

# ggplot
#library(ggplot2)
```

Stats for Manuscript
-----------------------------
Quantitative descriptions of composition and changes, and land use percentages, used in the manuscript

### Composition

```{r compStatsMS}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

pls <- read.csv("R output/OHCompCleanPLS_percent.csv")
fia <- read.csv("R Output/OHCompCleanFIA_percent.csv")

# assign county groups (1 = nw, 2 = se)
pls$region <- c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,2,1,2,2,1,2,2,2,2,2,2,2,2)
pls$regionName <- apply(as.array(pls$region),1,function (x) {
  if (x == 1) out <- 'Northwest'
  else out <- 'Southeast'
  return(out) })

fia$region <- c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,2,1,2,2,1,2,2,2,2,2,2,2,2)
fia$regionName <- apply(as.array(fia$region),1,function (x) {
  if (x == 1) out <- 'Northwest'
  else out <- 'Southeast'
  return(out) })

# split counties by region for summary stats
northwestPLS <- pls[pls$regionName == 'Northwest',]
southeastPLS <- pls[pls$regionName == 'Southeast',]

northwestFIA <- fia[fia$regionName == 'Northwest',]
southeastFIA <- fia[fia$regionName == 'Southeast',]

# mean and stdev for each region
nwPLS <- rbind(mean = colMeans(northwestPLS[,3:17]), sd = apply(northwestPLS[,3:17], 2, sd))
nwFIA <- rbind(mean = colMeans(northwestFIA[,3:17]), sd = apply(northwestFIA[,3:17], 2, sd))

sePLS <- rbind(mean = colMeans(southeastPLS[,3:17]), sd = apply(southeastPLS[,3:17], 2, sd))
seFIA <- rbind(mean = colMeans(southeastFIA[,3:17]), sd = apply(southeastFIA[,3:17], 2, sd))

# stats
nwPLS 
northwestPLS[,c('countyName','beech','oak','ash','elm','maple','hickory')]

nwFIA
northwestFIA[,c('County','beech','oak','ash','elm','maple','hickory')]

# change in proportion 
nwFIA['mean',] - nwPLS['mean',]


sePLS

seFIA

# change in proportion 
seFIA['mean',] - sePLS['mean',]

```


### Land Use
Summarize forest coverage in both regions

```{r landUseNumbersMS}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# read in data
landuse <- read.csv("R Output/OHCountyLandCover_aggregated_nmds.csv")
centroids <- read.csv("R output/OH_county_centroids_latlong.csv") 

# subset enviro and land use to 27 data counties, add coords
landuse2 <- merge(centroids, landuse, by.x = "ID", by.y = "Polygon", all.y = FALSE)

# 1 = northwest, 2 = southeast
landuse2$region <- c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,2,1,2,2,1,2,2,2,2,2,2,2,2)

regionSplit <- split(landuse2, landuse2$region)
names(regionSplit) <- c('Northwest','Southeast')

# Regional mean coverage by land use type
sapply(regionSplit, function(x) colMeans(x[,5:14]))
```

### Climate
Summarize climate in both regions

```{r ClimateStatsMS}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# read in data
envar <- read.csv("R output/OH_EnviroVarCountySummary.csv")
centroids <- read.csv("R output/OH_county_centroids_latlong.csv") 

# subset enviro and land use to 27 data counties, add coords
envar2   <- merge(centroids, envar, by.x = "ID", by.y = "countyID", all.y = FALSE)

# 1 = northwest, 2 = southeast
envar2$region <- c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,2,1,2,2,1,2,2,2,2,2,2,2,2)

regionSplit <- split(envar2, envar2$region)
names(regionSplit) <- c('Northwest','Southeast')

# Regional mean coverage by land use type
sapply(regionSplit, function(x) colMeans(x[,5:11]))

```

