```{r options, include=FALSE}
#opts_chunk$set(cache = TRUE)
```

Ohio PLS and FIA Data Processing
========================================================

Jillian Deines  
March 2014  

This script processes the PLS GIS data shapefile and FIA summaries (crunched by Dan Williams) into aggregate species groups and county percentages. It also creates the pie chart maps and change in species percentage figure.

Previous versions of this (ie, Ohio.PLS.FIA.landUseSummariesV6_2014.r) also summarized composition by township, but this is not reproduced here for streamlining. Original code also has contributions from Dan Williams.

#### Inputs:
* OHcorners.shp - GIS file used for data collection from georeferenced county maps
* OHcounty.shp  - Ohio county shapefile 
* FIASpringSpeciesPercentage.csv - FIA data aggregated by Dan Williams

#### Crunching:
* Summarizes PLS raw species counts by county
* Data Cleaning and Aggregation: PLS
* Data Cleaning and Aggregation: FIA
* Figure: Change in Species Percentage

#### Outputs:
* OHCountyCompRaw_PLS.csv - raw PLS species counts by county 
* OHCompClean_PLS.csv
* OHCompCleanPLS_percent.csv
* OHCompCleanFIA_percent.csv
* Figure: PLS vs FIA pie charts
  * OH_county_centroids_projected.csv
  * OH_county_centroids_latlong.csv
* Figure: Change in species percentage (SpeciesChange)

#### Next steps:
* Processing environmental and land use covariates in OH_Land_Use_Enviro_Var_processing.Rmd
* Run analyses in OHAnalyses.Rmd

Working directory and packages
------------------------------
This assumes you have the file 'CleanCodeAndData' set as your working directory

```{r setup,message=FALSE}
library(rgdal)
library(maptools)
library(PBSmapping)
library(plotrix)
library(plyr)
```


PLS Processing: Summarize Raw Data by County
---------------------------------------

```{r PLStoCountyRaw}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# load corner point shapefile (data collected from georeferenced maps). Current version OHcorners.shp output based: January 30, 2013. CRS = NAD83, latlong
corners <- readOGR("GISlayers","OHcorners", verbose=F)

# county shapefile
county <- readOGR("GISlayers","OHCounty", verbose=F) 

# quick look at data
par(mar=c(0,0,0,0))
plot(county, main='PLS Data Points')
plot(corners, col="blue", cex = 0.4,add=T)

# general data summary
numDataPoints <- length(corners)
numTrees <- sum(!is.na(corners$SP1)) + sum(!is.na(corners$SP2)) + sum(!is.na(corners$SP3)) + sum(!is.na(corners$SP4))

# make a vector of species possibilities = all species recorded from Ohio maps
treeNames <- c(levels(corners$SP1), levels(corners$SP2), levels(corners$SP3), levels(corners$SP4))
species <- sort(unique(treeNames)) # alphabetize
species <- species[!species %in% c('no tree')] # remove 'no tree' category

# Identify which polygon (in "county") each point in corners falls within 
corners$countyID <- over(corners, county)$ID


# Calculate total # of each species in each county and store it as matrix "comp"
comp <- t(sapply(as.numeric(levels(factor(corners$countyID))), function(j)  {
	        J <- corners[ !is.na(corners$countyID) & corners$countyID==j,]  
	        c(county=j, sapply(species, function(i) {
            sum(c(J$SP1==i,J$SP2==i,J$SP3==i,J$SP4==i),na.rm=TRUE)
            }))
          }
        ))

# sum the trees for each county (sum row excluding county column)
comp.df <- as.data.frame(comp)
comp.df$total <- rowSums(comp.df[,-1])

# add county names to output
countyKey <- as.data.frame(county[,11:12])
comp.df2 <- merge(comp.df,countyKey,by.x='county', by.y='ID')
comp.df2 <- comp.df2[,c(1,73,2:72)]

# replace spaces from column names with period
colnames(comp.df2) <- gsub("\\ ", ".", colnames(comp.df2)) 

# write output to CSV for data sharing
write.csv(comp.df2,"R output/OHCountyCompRaw_PLS.csv", row.names=F)
```

This Ohio dataset includes `r numDataPoints` corner data locations and `r numTrees` trees recorded. There were `r length(species)` names recorded in the PLS dataset.  

Data Cleaning and Aggregation: PLS
-----------------------------------
This aggregates the `r length(species)` common names into 15 taxonomic groups. Next, it excludes counties with low sample numbers from analysis (because the counties don't directly align with the township borders, there are a few counties with a small number of points that belong to townships mostly within neighboring counties).  

```{r cleanPLS, fig.height=5, fig.width=5}
raw.comp <- comp.df2

#----PLS taxon groupings----# 
raw.comp2 <- cbind(raw.comp[,c('county','CountyName')],
	"ash"=rowSums(raw.comp[,c("ash", "black.ash", "blue.ash", 
                             "hoop.ash", "swamp.ash", "white.ash")]),
	"basswood"=rowSums(raw.comp[,c("basswood","linden","lynn")]),
	"beech"=raw.comp$beech,
	"cherry"=raw.comp$cherry,
	"elm"=rowSums(raw.comp[,c("elm", "red.elm")]),
	"hackberry"=raw.comp$hackberry,
	"hickory"=rowSums(raw.comp[,c("bitternut", "hickory")]),
	"honeylocust"=raw.comp$honey.locust,
	"maple"=rowSums(raw.comp[,c("maple", "silver.maple", "soft.maple", "sugar",
                  "sugar.maple", "white.maple", "box.elder.maple", "box.elder")]),
	"oak"=rowSums(raw.comp[,c("black.jack", "black.oak", "bur.oak", "chestnut.oak",
                "jack.oak", "oak", "pin.oak", "post.oak", "red.oak", "spanish.oak",
                "shingle.oak", "swamp.oak", "white.oak", "yellow.oak")]),
	"pine"=rowSums(raw.comp[,c("pine", "pitch.pine")]),
	"sycamore"=raw.comp$sycamore,
	"tulip"=raw.comp$poplar,
	"walnut"=rowSums(raw.comp[,c("black.walnut", "walnut", "white.walnut")]),
	"other"=rowSums(raw.comp[,c("red.bud", "chestnut", "hemlock", 
              "birch", "aspen", "cottonwood", 'persimmon', "trembling.aspen", 
              "gum", "black.gum", "dogwood", "pepperidge", "mulberry", "ironwood",
              "serviceberry", "hornbeam", "coffeetree", "cedar", "locust", 
              "black.locust", "buckeye","sassafras", "willow", "sourwood")]),
	"total"=raw.comp$total)

# check to make sure we got all species by re-summing the total count for each
# county in the combined taxons and checking against the original total 
test.comp2 <-cbind(raw.comp2, "check.total"=rowSums(raw.comp2[,3:17])) #re-sum trees
sum(test.comp2$total != test.comp2$check.total) 


#----CLEAN - REMOVE COUNTIES WITH LOW NUMBERS----#

# view distribution of tree counts by county
plot(raw.comp2$total, ylab="Trees per County", xlab='County Index')  # clear break below 262
abline(h=240, col='red')

# remove all rows with a total tree count less than 262 (27 counties left)
pls.comp <- subset(raw.comp2, raw.comp2$total > 261)

# see how many trees per county
summary(pls.comp$total)

# get new tally of remaining trees in the dataset
sum(pls.comp$total)

#----CONVERT SPECIES NUMBERS TO PERCENTAGES ##

# divide each cell with tree counts by the total tree count
pls.comppct <- cbind(countyID = pls.comp$county, countyName = pls.comp$CountyName, 
                     pls.comp[,3:17]/pls.comp$total, total = pls.comp$total)

setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')
write.csv(pls.comp,"R output/OHCompClean_PLS.csv",row.names=F)
write.csv(pls.comppct, "R output/OHCompCleanPLS_percent.csv", row.names=F)
```

Data Cleaning and Aggregation: FIA
-----------------------------------------
Dan Williams downloaded data from FIA. He specified all species/individuals among a certain size class (> 9 in diameter) to match PLS methodology. Dan converted raw tree counts into percentage data in Excel. PDF's of downloaded county data are in the folder "FIA Tree Count PDFs", and I am guessing intermediate processing steps are in the folder 'FIA Ohio Summarized Data'. Both of these files are located in the 'FIAdataFromDan' folder in the working directory (CleanCodeAndData).

May 14, 2014 - Revised so 'other' count does not include dead trees ('unknown dead hardwood' and 'unknown dead conifer')

```{r cleanFIA}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# read in species level FIA county percentage data compiled by Dan
FIA.spec1 <- read.csv("FIAdataFromDan/FIASpringSpeciesPercentage.csv")

# check PLS and FIA counties match
sum(sort(pls.comp$county) != sort(FIA.spec1$id))

#----Combine Species Classes----#
FIA.spec2 <-cbind(FIA.spec1[,c("id","County")],
	"ash" = rowSums(FIA.spec1[,c("ash", "white.ash", "black.ash", "green.ash", 
                               "blue.ash")]),
	"basswood" = FIA.spec1$american.basswood,
	"beech"= FIA.spec1$american.beech,
	"cherry" = rowSums(FIA.spec1[,c("black.cherry","pin.cherry","chokecherry",
                                  "sweet.cherry")]),
	"elm" = rowSums(FIA.spec1[,c("american.elm", "siberian.elm", "slippery.elm",
                               "rock.elm")]),
	"hackberry" = FIA.spec1$hackberry,
	"hickory" = rowSums(FIA.spec1[,c("hickory", "bitternut.hickory", "pignut.hickory", 
                    "shellbark.hickory", "shagbark.hickory","mockernut.hickory")]),
	"honeylocust" = FIA.spec1$honeylocust,
	"maple" = rowSums(FIA.spec1[,c("black.maple", "red.maple", "silver.maple", 
                                 "sugar.maple", "norway.maple", "boxelder")]),
	"oak"=rowSums(FIA.spec1[,c("oak", "white.oak", "swamp.white.oak", "scarlet.oak",
                "northern.pin.oak", "shingle.oak", "overcup.oak", "bur.oak", 
                "swamp.chestnut.oak", "chinkapin.oak", "pin.oak", "chestnut.oak", 
                "northern.red.oak", "post.oak", "black.oak")]),
	"pine" = rowSums(FIA.spec1[,c("eastern.white.pine", "scotch.pine", 
                                "loblolly.pine", "virginia.pine", "austrian.pine",
                                "red.pine", "pitch.pine", "shortleaf.pine")]),
	"sycamore" = FIA.spec1$american.sycamore,
	"tulip" = FIA.spec1$yellow.poplar,
	"walnut" = rowSums(FIA.spec1[,c("black.walnut","butternut")]),
	"other" = rowSums(FIA.spec1[,c("chestnut","eastern.red.cedar","norway.spruce",
                  "blue.spruce", "eastern.hemlock","ohio.buckeye","yellow.buckeye",
                  "ailanthus","european.alder","yellow.birch","sweet.birch",
                  "river.birch", "northern.catalpa","eastern.redbud","hawthorn",
                  "common.persimmon", "kentucky.coffeetree","sweetgum",
                  "osage.orange","cucumbertree","apple", "white.mulberry",
                  "blackgum","eastern.hophornbeam","sourwood",
                  "paulownia..empress.tree.","eastern.cottonwood","bigtooth.aspen",
                  "quaking.aspen","black.locust","black.willow","sassafras",
                  "other.or.unknown.live.tree")]),
	"total"=rowSums(FIA.spec1[,3:93]))

# sort data frame so rows increase by county id number
FIA.spec2 <- FIA.spec2[order(FIA.spec2$id),]  			# sort rows by county number

write.csv(FIA.spec2, "R Output/OHCompCleanFIA_percent.csv",row.names=F)
```

Figure: PLS vs FIA pie charts
-------------------------------
This creates MS figure 3
```{r pieChart, fig.width=8,fig.height=6}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# fia data
fia <- read.csv("R Output/OHCompCleanFIA_percent.csv")
# pls data
pls <- read.csv("R output/OHCompCleanPLS_percent.csv")

# read in county shapefile and reproject for figure using
# Ohio state plane coordinates, NAD83/Ohio South, EPSG 32123
county <- readOGR("GISlayers","OHCounty", verbose=F) 
county.proj <- spTransform(county, CRS("+init=epsg:32123"))

# Join data with county shapefile by the county ID number ("ID" in plscomp, "ID" in county)
pls2 <- merge(county.proj, pls, by.x= "ID", by.y= "countyID", all.x=F)
fia2 <- merge(county.proj, fia, by.x= "ID", by.y= "id", all.x=F)
latlong <- merge(county, pls, by.x= "ID", by.y= "countyID", all.x=F)

## calculate polygon centroids for 27 data counties
pls.polyset <- SpatialPolygons2PolySet(pls2)
latlong.polyset <- SpatialPolygons2PolySet(latlong)
centroids <- calcCentroid(pls.polyset, rollup=1)
centroidsLatLong <- calcCentroid(latlong.polyset, rollup=1)


# combine centroids with data
pls3 <- cbind(centroids, pls2)
fia3 <- cbind(centroids, fia2)
ll   <- cbind(centroidsLatLong, latlong)

#write.csv(pls3[,c('ID','X','Y')], "R output/OH_county_centroids_projected.csv", row.names=F)
#write.csv(ll[,c('ID','X','Y')], "R output/OH_county_centroids_latlong.csv", row.names=F)

# extract only necessary columns
speciesList <- names(pls3[17:31])
pls.pies <- pls3[,c('ID','X','Y',speciesList)]
fia.pies <- fia3[,c('ID','X','Y',speciesList)]

#----plot the pies!----

# set colors manually, using the 9 colors in "Set1" of RColorBrewer and then adding my own
col <- c("#377EB8", "#984EA3", "#4DAF4A", "#E41A1C", "cyan2", "green", "#F781BF",
         "#0C2C84", "#FFFF33", "#FF7F00", "#A65628", "olivedrab1", "#999999",
         "#E7298A", "white")
# legend names (corresponds with color order)
species <- c("ash", "basswood", "beech", "cherry", "elm", "hackberry", "hickory", 
             "honey locust", "maple", "oak", "pine", "sycamore", "tulip poplar", 
             "walnut", "other")

# turn on to write file to disk
#png("exportedFigs/PieCharts.png", type = "cairo", units = "in", width = 7, height = 4.5, res=2000)

par(mfrow=c(1,2))

# PLS
par(mar = c(2, 0.1, 0, 0.2)) 
plot(county.proj, col='azure3',border='white',lwd=2) 
#title(main="Settlement Era")
# add pie chart
for (i in 1:27) {
  floating.pie(pls.pies[i,2], pls.pies[i,3], 
               x=pls.pies[i,4:18][!pls.pies[i,4:18]==0], radius=14000, 
               col=col[!pls.pies[i,4:18]==0], edges=500)
}


# FIA
par(mar = c(2, 0.2, 0, 0.1)) 
plot(county.proj, col='azure3',border='white',lwd=2) 
#title(main="Modern")
# add pie chart
for (i in 1:27) {
  floating.pie(fia.pies[i,2], fia.pies[i,3], 
               x=fia.pies[i,4:18][!fia.pies[i,4:18]==0], radius=14000, 
               col=col[!fia.pies[i,4:18]==0], edges=500)
}

par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), mar=c(0, 0, 0, 0), new=TRUE)
plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
# add legend
legend("bottom", species, fill = col, border = "black", ncol=5, bty='n')
text(.6,.93,"Modern", cex=1.2)
text(-.5,.93,"Pre-settlement", cex=1.2)

#dev.off()

```

Figure: Change in Species Percentage
-----------------------------------

```{r SpeciesChange, fig.width=9, fig.height=6}
setwd('C:/Users/deinesji/Dropbox/Dan_Williams/CleanCodeAndData')

# fia data
fia <- read.csv("R Output/OHCompCleanFIA_percent.csv")
# pls data
pls <- read.csv("R output/OHCompCleanPLS_percent.csv")

# calculate difference
change2 <- cbind(fia[,c('id','County')], fia[,3:17]-pls[,3:17])
# convert proportional data to percentage
change2[,3:17] <- change2[,3:17]*100

# read in county shapefile and reproject for figure using
# Ohio state plane coordinates, NAD83/Ohio South, EPSG 32123
county <- readOGR("GISlayers","OHCounty", verbose=F) 
county.proj <- spTransform(county, CRS("+init=epsg:32123"))

# Join data with county shapefile by the county ID number 
change3 <- merge(county.proj, change2, by.x= "ID", by.y= "id", all.x=TRUE)

#png("exportedFigs/SpeciesChange_aw.png", type = "cairo", units = "in", width = 7, height = 5, res = 2000)

# alternative color scheme
#library(RColorBrewer) # used to get 'Spectral' color scheme, which I modifed below
library(grid)
labels <- names(change3[,14:28])
labels[c(8,13)] <- c('honey locust', 'tulip poplar')

colchange <- c("#a50026","#d73027","#f46d43","#fdae61","#fed569","antiquewhite3",
               "#abd9e9","#92c5de","#74add1","#4575b4","#313695")

#colchange <- c("#a50026","#d73027","#f46d43","#fdae61","#fed569","darkgray",
#               "#abd9e9","#92c5de","#74add1","#4575b4","#313695")

spplot(change3[,14:28],col.regions=rev(colchange), col='gray29',
       names.attr=labels, at=seq(-55,55,(55+55)/11), as.table=T, 
       #colorkey=list(labels=list(cex=1.2)), 
       par.settings = list(panel.background=list(col="white"),
                           layout.widths=list(right.padding=6)))
# legend title
grid.text('Change in Relative Abundance (%)',x=unit(0.95, "npc"),y=unit(0.5, 'npc'), rot=-90)

#dev.off()

```




